All just Greedy for Money? - A prediction and Analysis of the interplay between Graduate Statistics and Salary Prospects in the STEM Field

### 2024-01-28 doc/DataLiteracyStudentProject/paper.tex
update

Our central contribution is a Vector autoregressive model (VAR) discussed in \Cref{sec:methods} that is built using just mentioned data. Combined with a
Partial autocorrelation function (PACF) the optimal value for lag
is determined, see \Cref{sec:methods} \textcolor{red}{subsection}, to ensure that relevant structure in the data is captured by the model.
After verifying that the VAR model functions as expected and testing the model on a test set, \Cref{sec:methods} \textcolor{red}{subsection}, we
can predict all three input time series. As we are only interested in the number of students enrolled and the development of the salary, our research is 
focused on these two instances.
This allows us to conclude that salary has a significant impact on the amount of STEM students. Further, we can model different salary development scenarios, 
and see how the students would likely react (\Cref{sec:results}).

@@

Each data set used in the analysis had to be processed, to work for our analysis.
The university data is only available in PDF format to download. To process the data, it had
to be transformed into a more versatile format, using two steps. First, it has been transformed into CSV, which resulted in a corrupted file.
A custom-written parser was then used, to restore the original data structure.
% I am currently writing here

Our main goal was to find a correlation between the salary, the graduates and the resulting amount of students enrolling a subject in the STEM field. For this problem we use a vector autoregressive model, which predicts a time series 
for each input time series. The model creates a linear combination for each prediction which could be expressed in a matrix-vector notation.
\begin{equation*}
    \resizebox{\linewidth}{!}{%
        $\begin{bmatrix}
            S_{1,n}\\
            S_{2,n}\\
            \vdots \\
            S_{m,n}
        \end{bmatrix} = 
        \begin{bmatrix}
            C_{1,1} & C_{1,2} & \dots &C_{1,n-1}\\
            C_{2,1} & C_{2,2} &\dots & C_{2,n-1}\\
            \vdots & \vdots  & \ddots &\vdots\\
            C_{m,1} & C_{m,2} &\dots &C_{m,n-1}\\
        \end{bmatrix} \cdot
        \begin{bmatrix}
            S_{1,1} & S_{1,2} & \dots & S_{1,n-1}\\
            S_{2,1} & S_{2,2} & \dots & S_{2,n-1}\\
            \vdots & \vdots  & \ddots &\vdots\\
            S_{m,1} & S_{m,2} & \dots & S_{m,n-1}\\
        \end{bmatrix}+
        \begin{bmatrix}
            \epsilon_{1}\\
            \epsilon_{2}\\
            \vdots\\
            \epsilon_{n}\\
        \end{bmatrix}$%
    }
\end{equation*}
The $\epsilon$ vector approximates an error, which is induced by the model. One main benefit of the VAR model is that it also assumes that $S_{i,j}$ is also influenced by $S_{k,l}$ 
$i,k \in \{1,\dots, m\}: i\neq k$ and $j,l \in \{1,\dots, n-1\}: j\neq l$.
In our case, we have three input time series, the students, salary and graduates data from the Winter Semester 2007 to the Summer Semester 2021. This model has one hyperparameter which describes
the amount of past data that should be included in our models prediction. 
The vector autoregressive model heavily depends on the stationarity assumption which ensures that the stochastical process stays the same over time. 
In this context, this is assured by the following three axioms
\begin{enumerate}[label=\Roman*]
    \item The mean $\mu$ is constant
    \item The standard deviation is constant
    \item There is no seasonality in the data
\end{enumerate}
We assume that our time series is mostly stationary, and this is enough for our case because we are only predicting 2 data points in the future.
In order to find the optimal hyperparameter we used two well-established methods, the PACF and also an information-based criteria particularly the AIC and 
BIC. The PACF only considers the impact of the previous data point $S_{n-1}$ onto $S_{n}$, and then it builds a $(1-\alpha)$ confidence interval. We choose $\alpha = 5\%$ and get
get a 95\% Confidence Interval 
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 1.25723662 & 1.27163776 & 3.94417629 \\
        \hline
        Prediciton 2 & 4.84256257 & 0.11922267 & 1.72230067
    \end{tabular}
    \caption{Percentage difference between true value and predicted value for students, salary and graduates for lag 1}
\end{table}
As you can see, the percent difference between the true values especially for the students are quite high, becasue of that we tried the AIC and BIC to find a better lag for the 
hyperparameter p.  The abbreviation AIC stands for Akaike Information Criterion and tries to find an optimum between the complexity of the model and fitting the data, which should pretend overfitting. 
BIC is an abbreviation for Bayesian Information Criterion and the main difference between AIC and BIC is that BIC penalizes more complex models, which overall leads to less complicated models than the ones from AIC.
The optimal lag based on AIC is 6, while for BIC it is 1, which aligns with the PACF.
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 0.48413863\% & 0.41885829\% & 0.11117345\% \\
        \hline
        Prediciton 2 & 1.34596809\% & 0.30583828\% & 0.93859448\%
    \end{tabular}
    \caption{Percentage difference between true value and predicted value for students, salary and graduates for lag 5}
\end{table}
As you can see there is a major improvement in the prediction for the students in comparison to lag order 1. We decided to use lag 5, because the main focus of our 
of our research is to predict the students based on salary and graduates.
Another quite important aspect in the correlation analysis is the Granger Causality Test which we did on the output. 
Granger Causal implies that one time series is the result of another shifted time series.
In our case, we try to find out if e.g. the salary time series is just a shifted version of the students or the graduates time series.
The Granger causality test does exactly this, it uses the time series $S_i$ and tries to insert data points from $S_j$ or $S_k$ $i \neq j\neq k$
to predict the outcome of the model. After inserting the data points the test performs a t-test and a F-test, if both of these tests return statistically significant,
then we showed that $S_i$ could be expressed by a shifted input time series $S_j$ with $j\neq i$

>>>>

Our central contribution is a Vector autoregressive model (VAR) discussed in \Cref{sec:methods:var} that is built using just mentioned data. Combined with a
Partial autocorrelation function (PACF) the optimal value for lag
is determined, see \Cref{sec:methods:pacf}, to ensure that relevant structure in the data is captured by the model.
After verifying that the VAR model functions as expected and testing the model on a test set, \Cref{sec:methods} \textcolor{red}{subsection}, we
can predict all three input time series. As we are only interested in the number of students enrolled and the development of the salary, our research is 
focused on these two instances.
This allows us to conclude that salary has a significant impact on the amount of STEM students. Further, we can model different salary development scenarios, 
and see how the students would likely react (\Cref{sec:results}).

@@

\subsection{Data Processing}\label{sec:methods:processing}
Each data set used in the analysis had to be processed, to work for our analysis.
The university data is only available in PDF format to download. To process the data, it had
to be transformed into a more versatile format, using two steps. First, it has been transformed into CSV, which resulted in a corrupted file.
A custom-written parser was then used, to restore the original data structure.
In the year 2012, two years of high school graduates finished, due to the transition from G9 to G8 in
Baden-Württemberg. This outlier in the data had to be flattened out to ensure the VAR model could be fitted correctly.
The model uses the mean square error as a loss function, which makes outliers influence the data disproportionately.
Half the number of students from the year 2012 has been spread over the next four years to flatten the data.
The Deutsches Statistisches Bundesamt from which the average gross salary is obtained, does not account for inflation in its data set.
Since we need to maintain comparability between past years, the salary has been adjusted for a yearly cumulative inflation rate.

\subsection{Model creation}\label{sec:methods:var}
A vector autoregressive model is used to analyze relationships between multiple time series.
The model creates a linear combination for each prediction which can be expressed in a matrix-vector notation.
\begin{equation*}
    \resizebox{\linewidth}{!}{%
        $\begin{bmatrix}
            S_{1,n}\\
            S_{2,n}\\
            \vdots \\
            S_{m,n}
        \end{bmatrix} = 
        \begin{bmatrix}
            C_{1,1} & C_{1,2} & \dots &C_{1,n-1}\\
            C_{2,1} & C_{2,2} &\dots & C_{2,n-1}\\
            \vdots & \vdots  & \ddots &\vdots\\
            C_{m,1} & C_{m,2} &\dots &C_{m,n-1}\\
        \end{bmatrix} \cdot
        \begin{bmatrix}
            S_{1,1} & S_{1,2} & \dots & S_{1,n-1}\\
            S_{2,1} & S_{2,2} & \dots & S_{2,n-1}\\
            \vdots & \vdots  & \ddots &\vdots\\
            S_{m,1} & S_{m,2} & \dots & S_{m,n-1}\\
        \end{bmatrix}+
        \begin{bmatrix}
            \epsilon_{1}\\
            \epsilon_{2}\\
            \vdots\\
            \epsilon_{n}\\
        \end{bmatrix}$%
    }
\end{equation*}
The $\epsilon$ vector approximates an error, which is induced by the model. One main benefit of the VAR model is that it also assumes
that $S_{i,j}$ is also influenced by $S_{k,l}$ $i,k \in \{1,\dots, m\}: i\neq k$ and $j,l \in \{1,\dots, n-1\}: j\neq l$.
A VAR model has one major advantage, over other possible approaches, which is its simplicity. Therefore its results remain interpretable.
This model only uses one hyperparameter which describes the amount of past data that is used to make a prediction. This parameter is called the lag.
The vector autoregressive model heavily depends on the stationarity assumption which ensures that the stochastical process stays the same over time. 
In this context, this is assured by the following three axioms:
\begin{enumerate}[label=\Roman*]
    \item The mean $\mu$ is constant
    \item The standard deviation is constant
    \item There is no seasonality in the data
\end{enumerate}
Since we are only interested in predicting two time steps into the future, which is equivalent to one year,
we can safely assume that our data is stationary.

\subsection{PACF}\label{sec:methods:pacf}
To find the optimal hyperparameter we used two well-established methods, the PACF and information-based criteria. Particularly the AIC and 
BIC. The PACF only considers the impact of the previous data point $S_{n-1}$ on $S_{n}$ to build a $(1-\alpha)$ confidence interval.
We choose $\alpha = 5\%$ and get get a 95\% Confidence Interval 
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 1.257\% & 1.271\% & 3.944\% \\
        \hline
        Prediciton 2 & 4.843\% & 0.119\% & 1.722\%
    \end{tabular}
    \caption{Difference in percent between the true value and predicted value for students, salary, and graduates. Lag set to 1.}
\end{table}
The percentage differences between the true values especially for the students are quite high. To get better results
the AIC and BIC are used to find a better lag for the hyperparameter $p$ \textcolor{red}{is $p$ = lag?}. The abbreviation AIC
stands for Akaike Information Criterion and tries to find an optimum between the complexity of the model and fitting the data.
This should prevent overfitting. 
BIC is short for Bayesian Information Criterion and penalizes more complex models.
Therefore BIC leads to less complicated models than the ones from AIC, which sets the two apart.
The optimal lag based on AIC is 6, while for BIC it is 1, which is consistent with the PACF.
\begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c}
        & students & salary & graduates \\
        \hline
        Prediciton 1 & 0.484\% & 0.419\% & 0.111\% \\
        \hline
        Prediciton 2 & 1.346\% & 0.306\% & 0.939\%
    \end{tabular}
    \caption{Percentage difference between the true value and predicted value for students, salary, and graduates. Lag set to 5.}
\end{table}

% I am currently here

As you can see there is a major improvement in the prediction for the students in comparison to lag order 1. We decided to use lag 5, because the main focus of our 
of our research is to predict the students based on salary and graduates.
Another quite important aspect in the correlation analysis is the Granger Causality Test which we did on the output. 
Granger Causal implies that one time series is the result of another shifted time series.
In our case, we try to find out if e.g. the salary time series is just a shifted version of the students or the graduates time series.
The Granger causality test does exactly this, it uses the time series $S_i$ and tries to insert data points from $S_j$ or $S_k$ $i \neq j\neq k$
to predict the outcome of the model. After inserting the data points the test performs a t-test and a F-test, if both of these tests return statistically significant,
then we showed that $S_i$ could be expressed by a shifted input time series $S_j$ with $j\neq i$
